Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.}

### 2022-04-20 analysis/paper/shceof.tex
Render

\begin{abstract}
abstract
\\
\keywords{
         \and
    }

@@

We restrict this paper to the spring September-October-November (SON) trimester. This season experiences a maximum in tropical teleconnections (Cazes-Boezio, Robertson, and Mechoso 2003) and stratosphere-troposphere interactions (Lim, Hendon, and Thompson 2018).

In Section~\ref{methods} we describe the methods.
In Section~\ref{spatial} we analyse the spatial patterns of each complex EOF.
In Section~\ref{regressions} we study the spatial regressions with geopotential height, temperature and ozone anomalies.
In Section~\ref{psa} and \ref{sam} we analyse the relationship between cEOF2 the PSA and SAM modes.
In Section~\ref{tropical} we study tropical forcings that explain the variability of each cEOF.
Finally, in Section \ref{precipitation} we show the relationship between these modes of variability and precipitation and surface temperature anomalies in South America and Oceania.

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; Hersbach et al. (2019){]} for the period 1979 -- 2019.
Most of our analysis is restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 (Bell et al. 2020) to describe long-term trends.
We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK (Adams, Swartztrauber, and Sweet 1999) and we computed horizontal wave activity fluxes following Plumb (1985).
Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 (Huang et al. 2017) and precipitation monthly data from the CPC Merged Analysis of Precipitation (P. Xie and Arkin 1997), with a 2\degree and 2.5\degree resolution in latitude and longitude respectively.
The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.

@@

We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period (Baldwin and Thompson 2009).
We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by Campitelli, Díaz, and Vera (2022).
Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated, in agreement with Mo and Paegle (2001), as the third and fourth leading EOF of seasonal mean anomalies for 500-hPa geopotential heights at SH.

@@

\begin{figure}
\centering
\includegraphics{../figures/ceofs-1-1.pdf}
\caption{\label{fig:ceofs-1}Spatial patterns for the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa for the 1979 -- 2019 period. The shading (contours) corresponds to real (imaginary) part. Arbitrary units. In parenthesis, the proportion of variance (with respect to the zonal mean) explained.}
\end{figure}

@@

Given the similarity between the cEOF2 regression maps (Fig.~\ref{fig:eof2-regr-gh}) and documented PSAs we study the relationship between these modes.
Table~\ref{tab:psa-eof2} shows the correlations between the two PSA indices and the time series for Real and Imaginary phases of cEOF2.
As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong positive correlation between PSA1 and Imaginary cEOF2, and also between PSA2 and Real cEOF2.
On the other hand, there is no significant relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, so it is possible to make an association between its two phases and the two PSA modes.
The same phase election for cEOF2 that maximises the relationship between ENSO and Imaginary cEOF2, also maximises the association between cEOF2 components and PSA modes (not shown).

@@

In the case of Imaginary cEOF2, the correlation between SAM and Imaginary cEOF2 for the troposphere is associated to the asymmetric variability of the SAM.
Indeed, the Imaginary cEOF2 shares up to 92\% variance with the A-SAM and only 12\% at most with the S-SAM (Figure~\ref{fig:sam-eof-vertical}.b2).
Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the methodology proposed in this work is able to characterise the zonally asymmetric component of the SAM described previously by Campitelli, Díaz, and Vera (2022).

@@

The cEOF2 is not associated with the same tropical SSTs for all their phases.
Figure~\ref{fig:sst-psi-2}.d1 and d2 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
As a result, the correlation between the Real cEOF2 and ENSO is also not significant (0 (CI: -0.3 -- 0.3)).
Meanwhile, Rows a and c in Fig.~\ref{fig:sst-psi-2} show that the intermediate phases are still associated with significant SST anomalies over the Pacific Ocean, but with with slightly different location.
The 135º phase is associated with SST in the central Pacific (Fig.~\ref{fig:sst-psi-2}a.1), while the 45º phase is associated with SST in the eastern Pacific (Fig.~\ref{fig:sst-psi-2}c.1), which correspond roughly to the Central Pacific and Eastern Pacific ``flavours'' of ENSO, respectively (Kao and Yu 2009).
Both of these phases could be also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics.

@@

Over the continents, the Imaginary cEOF2 (Fig.~\ref{fig:pp-temp-2}b.1) is associated with positive temperatures in southern Australia and negative temperature in southern South America and the Antarctic Peninsula, that are a result of the wave train described before.

@@

Some results of our work contrast with those obtained by Irving and Simmonds (2016) when analysing the PSA.
They did not find neither a strong association between PSA-like variability and ENSO, nor a relationship between this variability and Australian precipitation.
Part of the disagreement could be due to the fact that we focus only on SON, which is the season when the teleconnection pattern is strongest, instead of the whole year.
Moreover, Irving and Simmonds (2016) use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might also be an issue.
Additionally, the disagreement could be related to the fact that Irving and Simmonds (2016) show weak relationship between ENSO and their PSA pattern, which explains partly the relationship between Australian rainfall and PSA-like variability.

@@

\leavevmode\vadjust pre{\hypertarget{ref-cazes-boezio2003a}{}}%
Cazes-Boezio, Gabriel, Andrew W. Robertson, and Carlos R. Mechoso. 2003. {``Seasonal {Dependence} of {ENSO Teleconnections} over {South America} and {Relationships} with {Precipitation} in {Uruguay}.''} \emph{Journal of Climate} 16 (8): 1159--76. \url{https://doi.org/10.1175/1520-0442(2003)16\%3C1159:SDOETO\%3E2.0.CO;2}.

>>>>

\begin{abstract}
asdfsdf
\\
\keywords{
        Southern Hemisphere circulation \and
        Teleconnections \and
        Pacific South American Mode \and
        Southern Annular Mode \and
        Stratosphere \and
    }

@@

We restrict this paper to the spring September-October-November (SON) trimester because this season experiences a maximum in tropical teleconnections over South America (Cazes-Boezio, Robertson, and Mechoso 2003) and Southern Hemisphere stratosphere-troposphere interactions (Lim, Hendon, and Thompson 2018).

In Section~\ref{methods} we describe the methods.
In Section~\ref{spatial} we analyse the spatial patterns of each complex EOF.
In Section~\ref{regressions} we study the spatial regressions with geopotential height, temperature and ozone anomalies.
In Section~\ref{psa} and \ref{sam} we analyse the relationship between cEOF2 the PSA and SAM modes.
In Section~\ref{tropical} we study tropical forcings that explain the variability of each cEOF.
In Section~\ref{precipitation} we show the relationship between these modes of variability and precipitation and surface temperature anomalies in South America and Oceania.
In Section~\ref{discussion} we compare our results with previous studies and discuss the benefits of our methodology.
In Section~\ref{conclusion} we summarise the conclusions.

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; Hersbach et al. (2019){]} for the period 1979 -- 2019.
Most of our analysis is restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 (Bell et al. 2020) to describe long-term trends.
We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK (Adams, Swartztrauber, and Sweet 1999) and we computed horizontal wave activity fluxes following Plumb (1985).
Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 (Huang et al. 2017) and precipitation monthly data from the CPC Merged Analysis of Precipitation (P. Xie and Arkin 1997), with a 2\degree and 2.5\degree horizontal resolution respectively.
The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.

@@

We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period (Baldwin and Thompson 2009).
We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by Campitelli, Díaz, and Vera (2022).
Seasonal indices of the PSA patterns (PSA1 and PSA2) were calculated, in agreement with Mo and Paegle (2001), as the third and fourth leading EOF of seasonal mean anomalies for 500-hPa geopotential heights at SH.

@@

\begin{figure}
\centering
\includegraphics{../figures/ceofs-1-1.pdf}
\caption{\label{fig:ceofs-1}Spatial patterns for the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa for the 1979 -- 2019 period. The shading (contours) corresponds to real (imaginary) part. Arbitrary units. The proportion of variance explained for each mode with respect to the zonal mean is indicated in parenthesis.}
\end{figure}

@@

Given the similarity between the cEOF2 regression maps (Fig.~\ref{fig:eof2-regr-gh}) and documented PSA patterns we study the relationship between these modes.
Table~\ref{tab:psa-eof2} shows the correlations between the two PSA indices and the time series for Real and Imaginary phases of cEOF2.
As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong positive correlation between PSA1 and Imaginary cEOF2, and also between PSA2 and Real cEOF2.
On the other hand, there is no significant relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, so it is possible to make an association between its two phases and the two PSA modes.
The same phase election for cEOF2 that maximises the relationship between ENSO and Imaginary cEOF2, also maximises the association between cEOF2 components and PSA modes (not shown).

@@

In the case of Imaginary cEOF2, the correlation between SAM and Imaginary cEOF2 for the troposphere is associated to the asymmetric variability of the SAM.
Indeed, the Imaginary cEOF2 shares up to 92\% variance with the A-SAM and only 12\% at most with the S-SAM (Figure~\ref{fig:sam-eof-vertical}.b2).
Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the modes obtained in this work are able to characterise the zonally asymmetric component of the SAM described previously by Campitelli, Díaz, and Vera (2022).

@@

The cEOF2 is not associated with the same tropical SSTs for all their phases.
Figure~\ref{fig:sst-psi-2}.d1 and d2 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
As a result, the correlation between the Real cEOF2 and ENSO is also not significant (0 (CI: -0.3 -- 0.3)).
Meanwhile, Rows a and c in Fig.~\ref{fig:sst-psi-2} show that the intermediate phases are still associated with significant SST regressed anomalies over the Pacific Ocean, but with with slightly different location.
The 135º phase is associated with SST in the central Pacific (Fig.~\ref{fig:sst-psi-2}a.1), while the 45º phase is associated with SST in the eastern Pacific (Fig.~\ref{fig:sst-psi-2}c.1), which correspond roughly to the Central Pacific and Eastern Pacific ``flavours'' of ENSO, respectively (Kao and Yu 2009).
Both of these phases could be also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics.

@@

Over the continents, the Imaginary cEOF2 (Fig.~\ref{fig:pp-temp-2}b.1) is associated with positive regressed temperature anomalies in southern Australia and negative regressed anomalies in southern South America and the Antarctic Peninsula, that are a result of the wave train described before.

@@

Some results of our work contrast with those obtained by Irving and Simmonds (2016) when analysing the PSA.
They did not find neither a strong association between PSA-like variability and ENSO, nor a relationship between this variability and Australian precipitation.
Part of the disagreement could be due to the fact that we focus our analysis only on SON, which is the season when the teleconnection patterns are strongest, instead of the whole year.
Moreover, Irving and Simmonds (2016) use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might be another source of disagreement.
Additionally, the discrepancy could be related to the fact that Irving and Simmonds (2016) show a weak relationship between ENSO and their PSA pattern, which explains partly the relationship between Australian rainfall and PSA-like variability.

@@

\leavevmode\vadjust pre{\hypertarget{ref-cazes-boezio2003}{}}%
Cazes-Boezio, Gabriel, Andrew W. Robertson, and Carlos R. Mechoso. 2003. {``Seasonal {Dependence} of {ENSO Teleconnections} over {South America} and {Relationships} with {Precipitation} in {Uruguay}.''} \emph{Journal of Climate} 16 (8): 1159--76. \url{https://doi.org/10.1175/1520-0442(2003)16\%3C1159:SDOETO\%3E2.0.CO;2}.
